Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate

### 2017-12-27 paper/Scientific Reports/main.tex
Added Scientific Reports version
Fixed problem with labels
Updated scientific reports .tex file with bibliography inside .tex file and added covertletterSciRep.tex

\documentclass{article}
%\usepackage{jair, theapa, rawfonts}

%\usepackage[utf8]{inputenc}
%\usepackage{fontspec}
\usepackage{xeCJK}
%\usepackage{armtex}

@@

\usepackage[numbers]{natbib}
\bibliographystyle{plainnat}

%\title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
%\author{\name Marcelo Prates \email morprates@inf.ufrgs.br \\
%\name Luis Lamb \email lamb@inf.ufrgs.br \\
%\addr Institute of Informatics \\
%Federal University of Rio Grande do Sul \\
%Porto Alegre, 91501-970 Brazil}

\title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}
\author{Marcelo Prates \and Luis Lamb}

\begin{document}

\maketitle

\begin{abstract}
Recently, there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of bias and prejudices. A large number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas. Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of \emph{gender} bias in AI. 
In this paper, we construct sentences with gender neutral pronouns in different languages 
and translate these sentences into English using the Google Translate API. 
The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate 
API computes for each sentence. We show that not only there is a tendency towards translating pronouns to
masculine, but also that this tendency is exaggerated for sentences about male dominated fields such as Computer Science. On the other hand, sentences about artistic fields are less likely to yield male defaults. This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of the people behind their training datasets -- ourselves.
\end{abstract}

@@

\begin{figure}[h]
	\centering
	\fbox{\includegraphics[width=\linewidth]{pictures/screenshot-gtranslate-hungarian}}
	\label{fig:screenshot-gtranslate-hungarian}
	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation. This screenshot from Google Translate shows how occupations from traditionally male-dominated fields such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
\end{figure}

\section{Motivation}

If automatic translation tools are indeed gender biased as Figure \ref{fig:screenshot-gtranslate-hungarian} seems to suggest, then possibly this phenomenon can be probed to yield many valuable insights. One could in principle group different professional occupations according to their broad category (artistic, scientific, industrial, etc.) and evaluate whether the proportion of male pronouns is more pronounced in some fields compared to others. We expect, for example, that most sentences about computer-related jobs will be translated with male pronouns, as a reflection of the gender asymmetry of this job market. We can also collect translation statistics among varied gender neutral languages to see whether different cultures make different assumptions about the gender of a subject.

@@

\begin{table}[H]
	\centering
	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
	\hline
	Language Family & Language & Pronominal Gender System & Supported & Tested \\ \hline \hline
	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 			& \checkmark	& \checkmark	\\
									& Tagalog 				& \xmark 			& \xmark		& \xmark		\\ \hline
	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Finnish 				& \xmark 			& \checkmark	& \checkmark	\\
									& Hungarian 			& \xmark 			& \checkmark 	& \checkmark	\\ \hline
	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Bengali 				& \xmark 			& \checkmark 	& \xmark		\\
									& English 				& \checkmark 		& \checkmark 	& \xmark 		\\
									& Persian 				& \checkmark 		& \checkmark 	& \xmark 		\\ \hline
	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 			& \xmark 		& \xmark 		\\
									& Nepali 				& \xmark 			& \checkmark	& \xmark		\\
									& Oriya 				& \xmark 			& \xmark 		& \xmark 		\\ \hline
	\multirow{10}{*}{} 				& Japanese 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Korean 				& O 				& \checkmark 	& \xmark		\\
									& Turkish 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Yoruba 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Basque 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Swahili 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Chinese 				& O 				& \checkmark 	& \checkmark	\\
									& Cantonese 			& \xmark 			& \xmark		& \xmark		\\
									& Pipil 				& \xmark 			& \xmark		& \xmark		\\
									& Quechuan 				& \xmark 			& \xmark		& \xmark		\\ \hline
	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 		& \checkmark 	& \xmark		\\
									& Ido 					& O 				& \xmark 		& \xmark 		\\
									\cline{2-2}
									& Lingua Franca Nova 	& \xmark 			& \xmark 		& \xmark 		\\
									\cline{2-2}
									& Interlingua 			& \xmark 			& \xmark 		& \xmark 		\\ \hline
	\end{tabular}
	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}. Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional). For the purposes of this work, we have decided to work only with languages lacking such a system and, as such, Persian and Esperanto have been discarded. Languages lacking support from Google Translate have been discarded.}
	\label{tab:gender-neutral-languages}
\end{table}

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|}
	\hline
	Category 		& \# Occupations 	\\ \hline \hline
	Artistic 		& $102$ 			\\ \hline
	Computer 		& $19$ 				\\ \hline
	Corporate 		& $50$ 				\\ \hline
	Dance 			& $9$ 				\\ \hline
	Film/Television & $26$ 				\\ \hline
	Healthcare 		& $88$ 				\\ \hline
	Industrial 		& $26$ 				\\ \hline
	Science 		& $50$ 				\\ \hline
	Service 		& $10$ 				\\ \hline
	Theatre 		& $52$ 				\\ \hline
	Writing 		& $29$ 				\\ \hline
	\hline
	Total			& $462$				\\ \hline
	\end{tabular}
	\caption{Selected occupations obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Lists_of_occupations}, grouped by category. We have selected a total of 462 occupations from $11$ distinct groups (Artistic, Science, Service, etc.).}
	\label{tab:occupations}
\end{table}

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|}
	\hline
	Language 	& Sentence template \\ \hline \hline
	Malay		& dia adalah $\langle occupation \rangle$ \\ \hline
	Estonian	& ta on $\langle occupation \rangle$ \\ \hline
	Finnish		& hän on $\langle occupation \rangle$ \\ \hline
	Hungarian	& ő $\langle occupation \rangle$ \\ \hline
	Armenian	& na $\langle occupation \rangle$ e \\ \hline
	Japanese	& $\langle occupation \rangle$ です\\ \hline
	Turkish		& o bir $\langle occupation \rangle$ \\ \hline
	Yoruba		& o jẹ $\langle occupation \rangle$ \\ \hline
	Basque		& $\langle occupation \rangle$ da \\ \hline
	Swahili		& yeye ni $\langle occupation \rangle$ \\ \hline
	Chinese		& ta $\langle occupation \rangle$ \\ \hline
	\end{tabular}
	\caption{Templates used to infer gender biases in the translation to the English language.}
	\label{tab:templates}
\end{table}

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|c|c|}
	\hline
	stagehands & author & neurologist \\ \hline
	screenwriter & animator & marketing director \\ \hline
	biochemist & endocrinologist & freelancer \\ \hline
	neurosurgeon & computer scientist & petrochemical engineer \\ \hline
	food stylist & cardiothoracic surgeon & property master \\ \hline
	literary editor & video editor & animation director \\ \hline
	house manager & chief administrative officer & arts administration \\ \hline
	actor & dialysis technician & family nurse practitioner \\ \hline
	psychologist & chief creative officer & flash developer \\ \hline
	scenic artist & producer & medical laboratory scientist \\ \hline
	\end{tabular}
	\caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of 462 different occupations.}
	\label{tab:occupations-examples}
\end{table}

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|}
	\hline
	happy 		& sad 		& right 	\\ \hline
	wrong 		& afraid	& brave 	\\ \hline
	smart		& dumb		& proud 	\\ \hline
	ashamed		& kind		& cruel 	\\ \hline
	envious		& loving	& hateful 	\\ \hline
	modest 		& arrogant	& guilty	\\ \hline
	innocent	& helpless	& shy		\\ \hline
	\end{tabular}
	\caption{Adjectives}
	\label{tab:adjectives}
\end{table}

@@

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Category & Female & Male & Neutral & Ratio & Total \\ \hline
	\hline
	Artistic        & 188 & 518 & 72 & 2.755 	& 918 \\ \hline
	Computer        & 7   & 125 & 12 & 17.857 	& 171 \\ \hline
	Corporate       & 36  & 340 & 23 & 9.444 	& 450 \\ \hline
	Dance           & 31  & 33  & 8  & 1.064 	& 81  \\ \hline
	Film-television & 54  & 125 & 18 & 2.315 	& 234 \\ \hline
	Healthcare      & 176 & 483 & 64 & 2.744 	& 792 \\ \hline
	Industrial      & 40  & 135 & 29 & 3.375   	& 234 \\ \hline
	Science         & 34  & 357 & 25 & 10.500	& 450 \\ \hline
	Service         & 5   & 63  & 10 & 12.600	& 90  \\ \hline
	Theatre         & 75  & 296 & 38 & 3.947 	& 477 \\ \hline
	Writing         & 41  & 148 & 30 & 3.610 	& 261 \\ \hline
	\hline
	Total           & 687 & 2623 & 329 & 3.818 & 4158 \\ \hline
	\end{tabular}
	\label{tab:gender-by-category}
	\caption{Number of female, male and neutral pronominal genders per occupation category in the translated sentences. The corresponding sex ratios (\# Male / \# Female) show just how much male defaults are prominent in male dominated fields such as computer science, with up to $\approx 18$ occurrences of male pronouns for each of a female one.}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{pictures/gender-by-category}
	\label{fig:gender-by-category}
	\caption{The distribution of pronominal genders in the translated sentences not only suggests a tendency towards male defaults but also reflects the structure of male dominated fields, with the proportion of male pronouns amounting to $73\%$ in computer related jobs and $76\%$ in corporate jobs respectively. Because Google Translate occasionally fails to translate a sentence, the bars for some categories fail to add up to $100\%$.}
\end{figure}

@@

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Language & Female & Male & Neutral & Ratio & Total \\ \hline
	\hline
	Malay     & 47  & 415  & 0   & 8.830 	& 462 \\ \hline
	Estonian  & 130 & 332  & 0   & 2.554 	& 462 \\ \hline
	Finnish   & 179 & 283  & 0   & 1.581 	& 462 \\ \hline
	Hungarian & 189 & 270  & 1   & 1.429 	& 462 \\ \hline
	Armenian  & 101 & 360  & 1   & 3.564 	& 462 \\ \hline
	Japanese  & 2   & 215  & 0   & 107.500  & 462 \\ \hline
	Turkish   & 22  & 394  & 43  & 17.909	& 462 \\ \hline
	Yoruba    & 12  & 304  & 131 & 25.333	& 462 \\ \hline
	Basque    & 5   & 50   & 153 & 10.000   & 462 \\ \hline
	Swahili   & 76  & 386  & 0   & 5.079 	& 462 \\ \hline
	Chinese   & 14  & 385  & 23  & 27.500   & 462 \\ \hline
	\hline
	Total     & 777 & 3394 & 352 & 4.368	& 4158   \\ \hline
	\end{tabular}
	\caption{Number of female, male and neutral pronominal genders per language in the translated sentences. The corresponding sex ratios (\# Male / \# Female) show just how much male defaults are prominent in some languages such as Chinese, with almost 30 male pronouns for each female one.}
	\label{tab:gender-by-language}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{pictures/gender-by-language}
	\caption{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.46\%$ and $2.98\%$ for Japanese and Chinese respectively. Once again not all bars add up to $100\%$ as Google Translate occasionally fails to translate sentences, particularly in Japanese and Basque. Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns.}
	\label{fig:gender-by-language}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{pictures/scatterplot-languages}
	\caption{Scatter plot of the translated sentences' statistics. Each point, corresponding to a single occupation, is positioned in the horizontal axis according to its sex ratio and in the vertical axis according to the frequency with which it is translated using gender neutral pronouns. Both quantities are averaged over all tested languages.}
	\label{fig:scatterplot-languages}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{pictures/histogram-female}
	\caption{Histogram of the distribution of female pronouns among different occupation categories, with a distinctive peak at 0. The corresponding histogram for male pronouns (see Figure \ref{fig:histogram-male}), by constrast, peaks at 9.}
	\label{fig:histogram-female}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{pictures/histogram-male}
	\caption{Histogram of the distribution of male pronouns among different occupation categories, with a distinctive peak at 9. In contrast with the corresponding histograms for female and gender neutral pronouns (see Figures \ref{fig:histogram-female} and \ref{fig:histogram-neutral} respectively) which exhibit inverse correlations, here one can see a tendency towards normally-distributed pronoun counts.}
	\label{fig:histogram-male}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{pictures/histogram-neutral}
	\caption{Following the behavior observed for female pronouns, the histogram for the distribution of gender neutral pronouns among different categories shows a tendency towards an inverse correlation, with a distinctive peak at 0.}
	\label{fig:histogram-neutral}
\end{figure}

@@

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Adjective 	& Female & Male & Neutral & Ratio & Total \\ \hline
	\hline
	Shy     	& 6  & 2   & 2  & 0.333 	& 12  \\ \hline
	Attractive 	& 4  & 2   & 4  & 0.500   	& 12  \\ \hline
	Happy   	& 5  & 3   & 2  & 0.600   	& 12  \\ \hline
	Kind    	& 4  & 3   & 1  & 0.750 	& 12  \\ \hline
	Ashamed 	& 4  & 5   & 1  & 1.250 	& 12  \\ \hline
	Smart   	& 2  & 5   & 3  & 2.500   	& 12  \\ \hline
	Envious 	& 2  & 6   & 1  & 3.000   	& 12  \\ \hline
	Sad     	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
	Loving  	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
	Helpless 	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
	Brave   	& 2  & 7   & 1  & 3.500   	& 12  \\ \hline
	Proud   	& 2  & 7   & 1  & 3.500   	& 12  \\ \hline
	Hateful 	& 1  & 5   & 3  & 5.000   	& 12  \\ \hline
	Dumb    	& 1  & 6   & 2  & 6.000   	& 12  \\ \hline
	Innocent 	& 1  & 8   & 1  & 8.000   	& 12  \\ \hline
	Right   	& 0  & 7   & 3  & -   		& 12  \\ \hline
	\hline
	Total   	& 43 & 129 & 41 & 3     	& 264 \\ \hline
	\end{tabular}
	\caption{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.}
	\label{tab:gender-by-adjective}
\end{table}

>>>>

\documentclass[fleqn,10pt]{wlscirep}

@@

%\usepackage[numbers]{natbib}
%\bibliographystyle{plainnat}

\title{Assessing Gender Bias in Machine Translation -- A Case Study with Google Translate}

\author[1,*]{Marcelo Prates}
\author[1,+]{Luis Lamb}
\affil[1]{Federal University of Rio Grande do Sul, Institute of Informatics, Porto Alegre, 91501-970, Brazil}

\affil[*]{morprates@inf.ufrgs.br}
\affil[+]{lamb@inf.ufrgs.br}

%\keywords{Keyword1, Keyword2, Keyword3}

\begin{abstract}
Recently, there has been a growing concern in academia, industrial research labs and the mainstream commercial media about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of bias and prejudices. A large number of Artificial Intelligence tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas. Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of \emph{gender} bias in AI. 
In this paper, we construct sentences with gender neutral pronouns in different languages 
and translate these sentences into English using the Google Translate API. 
The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate 
API computes for each sentence. We show that not only there is a tendency towards translating pronouns to
masculine, but also that this tendency is exaggerated for sentences about male dominated fields such as Computer Science. On the other hand, sentences about artistic fields are less likely to yield male defaults. This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of the people behind their training datasets -- ourselves.
\end{abstract}
\begin{document}

\flushbottom
\maketitle
% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
%
%  Click the title above to edit the author information and abstract
%
\thispagestyle{empty}

@@

\begin{figure}[h]
	\centering
	\fbox{\includegraphics[width=14cm]{pictures/screenshot-gtranslate-hungarian}}
	\caption{Translating sentences from a gender neutral language such as Hungarian to English provides a glimpse into the phenomenon of gender bias in machine translation. This screenshot from Google Translate shows how occupations from traditionally male-dominated fields such as scholar, engineer and CEO are interpreted as male, while occupations such as nurse, baker and wedding organizer are interpreted as female.}
	\label{fig:screenshot-gtranslate-hungarian}
\end{figure}
If automatic translation tools are indeed gender biased as Figure \ref{fig:screenshot-gtranslate-hungarian} seems to suggest, then possibly this phenomenon can be probed to yield many valuable insights. One could in principle group different professional occupations according to their broad category (artistic, scientific, industrial, etc.) and evaluate whether the proportion of male pronouns is more pronounced in some fields compared to others. We expect, for example, that most sentences about computer-related jobs will be translated with male pronouns, as a reflection of the gender asymmetry of this job market. We can also collect translation statistics among varied gender neutral languages to see whether different cultures make different assumptions about the gender of a subject.

@@

\begin{table}[H]
\begin{small}
	\centering
	\begin{tabular}{|c|m{2cm}|m{2cm}|c|c|}
	\hline
	Language Family & Language & Pronominal Gender System & Supported & Tested \\ \hline \hline
	\multirow{2}{*}{Austronesian} 	& Malay 				& \xmark 			& \checkmark	& \checkmark	\\
									& Tagalog 				& \xmark 			& \xmark		& \xmark		\\ \hline
	\multirow{3}{*}{Finno-Ugric} 	& Estonian 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Finnish 				& \xmark 			& \checkmark	& \checkmark	\\
									& Hungarian 			& \xmark 			& \checkmark 	& \checkmark	\\ \hline
	\multirow{4}{*}{Indo-European} 	& Armenian 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Bengali 				& \xmark 			& \checkmark 	& \xmark		\\
									& English 				& \checkmark 		& \checkmark 	& \xmark 		\\
									& Persian 				& \checkmark 		& \checkmark 	& \xmark 		\\ \hline
	\multirow{3}{*}{Indo-Aryan} 	& Maithili 				& \xmark 			& \xmark 		& \xmark 		\\
									& Nepali 				& \xmark 			& \checkmark	& \xmark		\\
									& Oriya 				& \xmark 			& \xmark 		& \xmark 		\\ \hline
	\multirow{10}{*}{} 				& Japanese 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Korean 				& O 				& \checkmark 	& \xmark		\\
									& Turkish 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Yoruba 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Basque 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Swahili 				& \xmark 			& \checkmark 	& \checkmark	\\
									& Chinese 				& O 				& \checkmark 	& \checkmark	\\
									& Cantonese 			& \xmark 			& \xmark		& \xmark		\\
									& Pipil 				& \xmark 			& \xmark		& \xmark		\\
									& Quechuan 				& \xmark 			& \xmark		& \xmark		\\ \hline
	\multirow{4}{*}{Constructed} 	& Esperanto 			& \checkmark 		& \checkmark 	& \xmark		\\
									& Ido 					& O 				& \xmark 		& \xmark 		\\
									\cline{2-2}
									& Lingua Franca Nova 	& \xmark 			& \xmark 		& \xmark 		\\
									\cline{2-2}
									& Interlingua 			& \xmark 			& \xmark 		& \xmark 		\\ \hline
	\end{tabular}
	\caption{Selected gender neutral languages obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Gender_neutrality_in_genderless_languages}. Languages are grouped according to language families and classified according to whether they exhibit pronominal gender system (\checkmark: yes, \xmark:~no, O: it is optional). For the purposes of this work, we have decided to work only with languages lacking such a system and, as such, Persian and Esperanto have been discarded. Languages lacking support from Google Translate have been discarded.}
	\label{tab:gender-neutral-languages}
	\end{small}
\end{table}

\begin{table}[H]
\begin{small}
	\centering
	\begin{tabular}{|c|c|}
	\hline
	Category 		& \# Occupations 	\\ \hline \hline
	Artistic 		& $102$ 			\\ \hline
	Computer 		& $19$ 				\\ \hline
	Corporate 		& $50$ 				\\ \hline
	Dance 			& $9$ 				\\ \hline
	Film/Television & $26$ 				\\ \hline
	Healthcare 		& $88$ 				\\ \hline
	Industrial 		& $26$ 				\\ \hline
	Science 		& $50$ 				\\ \hline
	Service 		& $10$ 				\\ \hline
	Theatre 		& $52$ 				\\ \hline
	Writing 		& $29$ 				\\ \hline
	\hline
	Total			& $462$				\\ \hline
	\end{tabular}
	\caption{Selected occupations obtained from the Wikipedia article \url{https://en.wikipedia.org/wiki/Lists_of_occupations}, grouped by category. We have selected a total of 462 occupations from $11$ distinct groups (Artistic, Science, Service, etc.).}
	\label{tab:occupations}
	\end{small}
\end{table}

\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|}
	\hline
	Language 	& Sentence template \\ \hline \hline
	Malay		& dia adalah $\langle occupation \rangle$ \\ \hline
	Estonian	& ta on $\langle occupation \rangle$ \\ \hline
	Finnish		& hän on $\langle occupation \rangle$ \\ \hline
	Hungarian	& ő $\langle occupation \rangle$ \\ \hline
	Armenian	& na $\langle occupation \rangle$ e \\ \hline
	Japanese	& $\langle occupation \rangle$ です\\ \hline
	Turkish		& o bir $\langle occupation \rangle$ \\ \hline
	Yoruba		& o jẹ $\langle occupation \rangle$ \\ \hline
	Basque		& $\langle occupation \rangle$ da \\ \hline
	Swahili		& yeye ni $\langle occupation \rangle$ \\ \hline
	Chinese		& ta $\langle occupation \rangle$ \\ \hline
	\end{tabular}
	\caption{Templates used to infer gender biases in the translation to the English language.}
	\label{tab:templates}
	}
\end{table}

\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|c|c|c|}
	\hline
	stagehands & author & neurologist \\ \hline
	screenwriter & animator & marketing director \\ \hline
	biochemist & endocrinologist & freelancer \\ \hline
	neurosurgeon & computer scientist & petrochemical engineer \\ \hline
	food stylist & cardiothoracic surgeon & property master \\ \hline
	literary editor & video editor & animation director \\ \hline
	house manager & chief administrative officer & arts administration \\ \hline
	actor & dialysis technician & family nurse practitioner \\ \hline
	psychologist & chief creative officer & flash developer \\ \hline
	scenic artist & producer & medical laboratory scientist \\ \hline
	\end{tabular}
	\caption{A randomly selected example subset of thirty occupations obtained from our dataset with a total of 462 different occupations.}
	\label{tab:occupations-examples}
	}
\end{table}

\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|c|}
	\hline
	happy 		& sad 		& right 	\\ \hline
	wrong 		& afraid	& brave 	\\ \hline
	smart		& dumb		& proud 	\\ \hline
	ashamed		& kind		& cruel 	\\ \hline
	envious		& loving	& hateful 	\\ \hline
	modest 		& arrogant	& guilty	\\ \hline
	innocent	& helpless	& shy		\\ \hline
	\end{tabular}
	\caption{Adjectives}
	\label{tab:adjectives}
	}
\end{table}

@@

\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Category & Female & Male & Neutral & Ratio & Total \\ \hline
	\hline
	Artistic        & 188 & 518 & 72 & 2.755 	& 918 \\ \hline
	Computer        & 7   & 125 & 12 & 17.857 	& 171 \\ \hline
	Corporate       & 36  & 340 & 23 & 9.444 	& 450 \\ \hline
	Dance           & 31  & 33  & 8  & 1.064 	& 81  \\ \hline
	Film-television & 54  & 125 & 18 & 2.315 	& 234 \\ \hline
	Healthcare      & 176 & 483 & 64 & 2.744 	& 792 \\ \hline
	Industrial      & 40  & 135 & 29 & 3.375   	& 234 \\ \hline
	Science         & 34  & 357 & 25 & 10.500	& 450 \\ \hline
	Service         & 5   & 63  & 10 & 12.600	& 90  \\ \hline
	Theatre         & 75  & 296 & 38 & 3.947 	& 477 \\ \hline
	Writing         & 41  & 148 & 30 & 3.610 	& 261 \\ \hline
	\hline
	Total           & 687 & 2623 & 329 & 3.818 & 4158 \\ \hline
	\end{tabular}
	\caption{Number of female, male and neutral pronominal genders per occupation category in the translated sentences. The corresponding sex ratios (\# Male / \# Female) show just how much male defaults are prominent in male dominated fields such as computer science, with up to $\approx 18$ occurrences of male pronouns for each of a female one.}
	\label{tab:gender-by-category}
	}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=10cm]{pictures/gender-by-category}
	\caption{The distribution of pronominal genders in the translated sentences not only suggests a tendency towards male defaults but also reflects the structure of male dominated fields, with the proportion of male pronouns amounting to $73\%$ in computer related jobs and $76\%$ in corporate jobs respectively. Because Google Translate occasionally fails to translate a sentence, the bars for some categories fail to add up to $100\%$.}
	\label{fig:gender-by-category}
\end{figure}

@@

\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Language & Female & Male & Neutral & Ratio & Total \\ \hline
	\hline
	Malay     & 47  & 415  & 0   & 8.830 	& 462 \\ \hline
	Estonian  & 130 & 332  & 0   & 2.554 	& 462 \\ \hline
	Finnish   & 179 & 283  & 0   & 1.581 	& 462 \\ \hline
	Hungarian & 189 & 270  & 1   & 1.429 	& 462 \\ \hline
	Armenian  & 101 & 360  & 1   & 3.564 	& 462 \\ \hline
	Japanese  & 2   & 215  & 0   & 107.500  & 462 \\ \hline
	Turkish   & 22  & 394  & 43  & 17.909	& 462 \\ \hline
	Yoruba    & 12  & 304  & 131 & 25.333	& 462 \\ \hline
	Basque    & 5   & 50   & 153 & 10.000   & 462 \\ \hline
	Swahili   & 76  & 386  & 0   & 5.079 	& 462 \\ \hline
	Chinese   & 14  & 385  & 23  & 27.500   & 462 \\ \hline
	\hline
	Total     & 777 & 3394 & 352 & 4.368	& 4158   \\ \hline
	\end{tabular}
	\caption{Number of female, male and neutral pronominal genders per language in the translated sentences. The corresponding sex ratios (\# Male / \# Female) show just how much male defaults are prominent in some languages such as Chinese, with almost 30 male pronouns for each female one.}
	\label{tab:gender-by-language}
	}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=10cm]{pictures/gender-by-language}
	\caption{The distribution of pronominal genders per language also suggests a tendency towards male defaults, with female pronouns reaching as low as $0.46\%$ and $2.98\%$ for Japanese and Chinese respectively. Once again not all bars add up to $100\%$ as Google Translate occasionally fails to translate sentences, particularly in Japanese and Basque. Among all tested languages, Basque was the only one to yield more gender neutral than male pronouns.}
	\label{fig:gender-by-language}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=10cm]{pictures/scatterplot-languages}
	\caption{Scatter plot of the translated sentences' statistics. Each point, corresponding to a single occupation, is positioned in the horizontal axis according to its sex ratio and in the vertical axis according to the frequency with which it is translated using gender neutral pronouns. Both quantities are averaged over all tested languages.}
	\label{fig:scatterplot-languages}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=10cm]{pictures/histogram-female}
	\caption{Histogram of the distribution of female pronouns among different occupation categories, with a distinctive peak at 0. The corresponding histogram for male pronouns (see Figure \ref{fig:histogram-male}), by constrast, peaks at 9.}
	\label{fig:histogram-female}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=10cm]{pictures/histogram-male}
	\caption{Histogram of the distribution of male pronouns among different occupation categories, with a distinctive peak at 9. In contrast with the corresponding histograms for female and gender neutral pronouns (see Figures \ref{fig:histogram-female} and \ref{fig:histogram-neutral} respectively) which exhibit inverse correlations, here one can see a tendency towards normally-distributed pronoun counts.}
	\label{fig:histogram-male}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=10cm]{pictures/histogram-neutral}
	\caption{Following the behavior observed for female pronouns, the histogram for the distribution of gender neutral pronouns among different categories shows a tendency towards an inverse correlation, with a distinctive peak at 0.}
	\label{fig:histogram-neutral}
\end{figure}

@@

\begin{table}[H]
\small{
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Adjective 	& Female & Male & Neutral & Ratio & Total \\ \hline
	\hline
	Shy     	& 6  & 2   & 2  & 0.333 	& 12  \\ \hline
	Attractive 	& 4  & 2   & 4  & 0.500   	& 12  \\ \hline
	Happy   	& 5  & 3   & 2  & 0.600   	& 12  \\ \hline
	Kind    	& 4  & 3   & 1  & 0.750 	& 12  \\ \hline
	Ashamed 	& 4  & 5   & 1  & 1.250 	& 12  \\ \hline
	Smart   	& 2  & 5   & 3  & 2.500   	& 12  \\ \hline
	Envious 	& 2  & 6   & 1  & 3.000   	& 12  \\ \hline
	Sad     	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
	Loving  	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
	Helpless 	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
	Brave   	& 2  & 7   & 1  & 3.500   	& 12  \\ \hline
	Proud   	& 2  & 7   & 1  & 3.500   	& 12  \\ \hline
	Hateful 	& 1  & 5   & 3  & 5.000   	& 12  \\ \hline
	Dumb    	& 1  & 6   & 2  & 6.000   	& 12  \\ \hline
	Innocent 	& 1  & 8   & 1  & 8.000   	& 12  \\ \hline
	Right   	& 0  & 7   & 3  & -   		& 12  \\ \hline
	\hline
	Total   	& 43 & 129 & 41 & 3     	& 264 \\ \hline
	\end{tabular}
	\caption{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.}
	\label{tab:gender-by-adjective}
	}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=10cm]{pictures/barplot-adjectives}
	\caption{The distribution of pronominal genders for each word in Table \ref{tab:adjectives} shows how stereotypical gender roles can play a part on the automatic translation of simple adjectives. One can see that adjectives such as \emph{shy} and \emph{attractive} are predominantly translated with female pronouns, while words like \emph{guilty}, \emph{innocent}, \emph{wrong}, \emph{right}, \emph{arrogant} are almost exclusively translated with male pronouns. Objective statements have a tendency towards male defaults, while statements concerning emotional states (\emph{shy}, \emph{happy}, \emph{kind}, \emph{ashamed}) amass at the other extreme of the sex ratio spectrum.}
	\label{fig:barplot-adjectives}
\end{figure}
